Staged Abstract Interpreters

### 2019-08-15 paper/60_cases.tex

The abstract addresses we introduced in Section \ref{sai} is simply identical to
the variable names, which is a context-insensitive schema.
Through tweaking the address allocation strategy, we can achieve various
context-sensitive analysis \cite{DBLP:conf/icfp/Gilray0M16}. In this section, we
demonstrate a calling-context-sensitive analysis, i.e., $k$-CFA-like analysis
\cite{DBLP:journals/jfp/HornM12}, in our staged abstract interpreter.

\begin{lstlisting}
  type Time = List[Expr]
  case class KCFAAddr(x: Ident, time: Time) extends Addr
\end{lstlisting}

The timestamp is essentially a finite list of expressions that tracks the $k$
recent calling contexts. The definition of abstract addresses @KCFAAddr@ is also
changed to contain a identifier and the time when it get allocated, meaning that
this address points to some values under such calling context. If $k$ is 0, we
obtain a monovariant analysis as demonstrated before; if $k > 0$, we obtain a
family of analysis with increasing precision. Note that when $k = 0$, the
address space is statically determined by the set of variables appeared in the
program, and we may exploit it to optimize the generated code. But when $k > 0$,
the address generation happens at the next stage, i.e., analysis time.

\begin{lstlisting}
  type AnsM[T] = RepReaderT[RepStateT[
                   RepStateT[RepSetReaderStateM[Cache, Cache, ?], Time, ?], //add Time with a RepStateT
                 Store, ?], Env, T]
  type Result = (Rep[Set[(Value, Store, Time)]], Rep[Cache])
\end{lstlisting}

We then integrate the timestamp into our moand stack by adding a staged
@StateT@ monad. The result type is passingly get updated to contain a set of
tuples, where the time is added.

\begin{lstlisting}
  def tick(e: Expr): AnsM[Unit] = for {
    τ <- get_time; u <- liftM(StateTMonad.put((e::τ).take($k$)))
  } yield u
  def alloc(x: String): AnsM[Addr] = for {
    τ <- get_time
  } yield unit(KCFAAddr(x, τ)) // we use unit to turn a current-stage constant to next-stage value
\end{lstlisting}

@@

A commonly used technique to improve the running time in control-flow analysis
is store-widening. As shown by \cite{Darais:2015:GTM:2814270.2814308,
DBLP:journals/pacmpl/DaraisLNH17}, store-widening analysis can be easily
implemented in monadic abstract interpreter by swaping the @StateT@ (for store)
and the @SetT@ (for nondeterminism). In this section, we show this optimization
is orthogonal to staging and we just need to swap the same \textit{staged}
version of monads. After the adjustment, we have the following monad stack:

\begin{lstlisting}
  type AnsM[T] = RepReaderT[RepSetT[RepStateT[
                   RepReaderT[RepStateT[IdM, Cache, ?], Cache, ?], Store, ?], ?], Env, T]
\end{lstlisting}

@@

Abstract garbage collection is a technique to reclaim unreachable addresses in
the store while doing the abstract interpretation
\cite{Might:2006:IFA:1159803.1159807}. \citet{DBLP:journals/pacmpl/DaraisLNH17}
shows that, for big-step monadic abstract interpreter, we may add a @Reader@
monad to track a set of root addresses and compute the reachable addresses every
time when we obtain a value. Here we adopt it with staged monads:

\begin{lstlisting}
  type AnsM[T] = RepReaderT[RepReaderT[
                   RepStateT[RepSetReaderStateM[Cache, Cache, ?], Store, ?],
                 Set[Addr], ?], Env, T] //add a reader monad for Set[Addr]
\end{lstlisting}

@@

\subsection{Modular Analysis for Free} \label{modular}

@@

One of the challenges of modern static analysis is program usually depends on
large libraries programs~\cite{toman_et_al:LIPIcs:2017:7121}. If we can analyze
programs and libraries separately and reuse the result without losing precision,
then we can reduce the overhead of repeatedly analyzing libraries.
Some static analyzers compute a summary for a function or a module, which
can be reused later, however they are mostly too conservative or unsound, which
both lead to imprecision. In this section, we show meta-programming can be used
to create sound summaries in the code generation from, such generated code can
be resued later. The summary contains information that not depend on the
function argument.

As we see when compiling a higher-order function (closure), we specialize the
abstract interpreter to the body expression of that lambda term under some
environment, without knowing the actual arguments to the function. We may reuse
the infrastratures and refactor the @close@ function mentioned in Section
\ref{sai} to achieve modular analysis. The @close@ function takes a lambda term
@f@ and an environment, produces a compiled closure, which can be freely used at
the next stage.

\begin{lstlisting}
  def close(ev: Eval => Ans)(f: Expr, ρ: Rep[Env]): Rep[CompiledClo] = { ...}
\end{lstlisting}

\paragraph{Partially-Static Context} Under a polyvariant analysis (like
$k$-CFA), such summary generated by specializing static analyzer even could do
more for calling context by partially-static data optimization. For example,
consider that we have a library consists of three functions @f@, @g@ and @h@,
where @f@ calls @g@ and @h@ internally and returns @y@ fianlly (we extend the
language to have multiple arguments). @g@ and @h@ do not depend on @f@. Now we
would like to modularlly specialize these functions, starting from @f@ with an
initial environment containing @f@, @g@ and @h@.

\begin{lstlisting}
  def f(a, b, c) = val x = g(a, b); val y = h(x, c);  y
\end{lstlisting}

Since $k$ is bounded to a fixed number, after the analysis running $k$ steps
inside of @f@, the calling context will purely depend on the static structure of
our module being analyzed.

@@

To evaluates an assignment, we first evaluate its right-hand side, and then put
the value into the slot where the address of @v@ points to. For simplicity, we
elect to make the value of the assignment be an @void()@ value.

\begin{lstlisting}
case Assign(x, e) =>
  val (v, $\sigma$_*) = ev(e, $\rho$, $\sigma$)
  (void(), put($\sigma$_*, get($\rho$, x), v))
\end{lstlisting}

To evaluates a @while@ loop statement, we evaluate the condition first. Then
similar to how we treat @branch0@, we have a generic @branch@ function but works
on boolean values. For the @true@ branch, we recursively call @ev@ on a newly
constructed expression @Seq(e, While(t, e))@ meaning that first evaluates @e@
and the repeat the loop. Otherwise, for the other branch, we simply return a
void value and current store.

\begin{lstlisting}
case While(t, e) =>
  val (tv, t$\sigma$) = ev(t, $\rho$, $\sigma$)
  branch(tv, ev(Seq(e, While(t, e)), $\rho$, t$\sigma$), (void(), t$\sigma$))
\end{lstlisting}

@@

Interval domain is a relative simple domain which consists of two numeric
fields, an upper bound and a lower bound. Here we annotate the upper bound and
lower bound fields to be of type @Rep[Double]@. The operations such as @+@ of
two intervals produce a new

\begin{lstlisting}
case class Interval(lb: Rep[Double], ub: Rep[Double]) {
  def +(that: Interval): Interval = that match {
    case Interval(lb_, ub_) => Interval(lb + lb_, ub + ub_)
  }
  ...
}
\end{lstlisting}

>>>>

The abstract addresses we introduced in Section \ref{sai} is simply identical to
the variable names, which is a context-insensitive schema.
Through tweaking the address allocation strategy, we can achieve various
context-sensitive analysis \cite{DBLP:conf/icfp/Gilray0M16}. In this section, we
demonstrate a calling-context-sensitive analysis, i.e., $k$-CFA-like analysis
\cite{DBLP:journals/jfp/HornM12}, in our staged abstract interpreter.
\begin{lstlisting}
  type Time = List[Expr]
  case class KCFAAddr(x: Ident, time: Time) extends Addr
\end{lstlisting}

The timestamp is essentially a finite list of expressions that tracks the $k$
recent calling contexts. The definition of abstract addresses @KCFAAddr@ is also
changed to contain a identifier and the time when it get allocated, meaning that
this address points to some values under such calling context. If $k$ is 0, we
obtain a monovariant analysis as demonstrated before; if $k > 0$, we obtain a
family of analysis with increasing precision. Note that when $k = 0$, the
address space is statically determined by the set of variables appeared in the
program, and we may exploit it to optimize the generated code. But when $k > 0$,
the address generation happens at the next stage, i.e., analysis time.
\begin{lstlisting}
  type AnsM[T] = RepReaderT[RepStateT[
                   RepStateT[RepSetReaderStateM[Cache, Cache, ?], Time, ?], //add Time with a RepStateT
                 Store, ?], Env, T]
  type Result = (Rep[Set[(Value, Store, Time)]], Rep[Cache])
\end{lstlisting}

We then integrate the timestamp into our moand stack by adding a staged
@StateT@ monad. The result type is passingly get updated to contain a set of
tuples, where the time is added.
\begin{lstlisting}
  def tick(e: Expr): AnsM[Unit] = for {
    τ <- get_time; u <- liftM(StateTMonad.put((e::τ).take($k$)))
  } yield u
  def alloc(x: String): AnsM[Addr] = for {
    τ <- get_time
  } yield unit(KCFAAddr(x, τ)) // we use unit to turn a current-stage constant to next-stage value
\end{lstlisting}

@@

A commonly used technique to improve the running time in control-flow analysis
is store-widening. As shown by \cite{Darais:2015:GTM:2814270.2814308,
DBLP:journals/pacmpl/DaraisLNH17}, store-widening analysis can be easily
implemented in monadic abstract interpreter by swaping the @StateT@ (for store)
and the @SetT@ (for nondeterminism). In this section, we show this optimization
is orthogonal to staging and we just need to swap the same \textit{staged}
version of monads. After the adjustment, we have the following monad stack:
\begin{lstlisting}
  type AnsM[T] = RepReaderT[RepSetT[RepStateT[
                   RepReaderT[RepStateT[IdM, Cache, ?], Cache, ?], Store, ?], ?], Env, T]
\end{lstlisting}

@@

Abstract garbage collection is a technique to reclaim unreachable addresses in
the store while doing the abstract interpretation
\cite{Might:2006:IFA:1159803.1159807}. \citet{DBLP:journals/pacmpl/DaraisLNH17}
shows that, for big-step monadic abstract interpreter, we may add a @Reader@
monad to track a set of root addresses and compute the reachable addresses every
time when we obtain a value. Here we adopt it with staged monads:
\begin{lstlisting}
  type AnsM[T] = RepReaderT[RepReaderT[
                   RepStateT[RepSetReaderStateM[Cache, Cache, ?], Store, ?],
                 Set[Addr], ?], Env, T] //add a reader monad for Set[Addr]
\end{lstlisting}

@@

\subsection{Modular Staged Analysis for Free} \label{modular}

@@

One of the challenges of modern static analysis is that program
usually depends on large libraries programs
\cite{toman_et_al:LIPIcs:2017:7121}. If we are able to analyze
programs and libraries separately and reuse the result without losing
precision, then we can reduce the overhead of repeatedly analyzing
libraries. Some static analyzers compute a summary for a function or a
module, which can be reused later. However, when lacking the context
information, those analyses can be too conservative or unsound, which
both lead to imprecision. In this section, we show that
meta-programming can provide a trade-off between summary-based modular
analysis and slow whole-program analysis, by ultilizing an existing
whole-program analyzer.

As we have already seen from Section \ref{sai}, and recall the
primitive operation @close@ that denotes a $\lambda$ term to next-stage
Scala function.  The specializing of staged abstract interpreter to
$\lambda$ terms is independent to other parts of the analyzed program.
\begin{lstlisting}
  def close(ev: Eval => Ans)(λ: Expr, ρ: Rep[Env]): Rep[Value]
\end{lstlisting}

Consider there is a library that contains a list map function @map@, 
\begin{lstlisting}
  (define (map xs f)
    (if (null? xs) '() (cons (f (car xs)) (map (cdr xs) f))))
\end{lstlisting}

The specialization (0CFA and without store-widening) of @close@ on
@map@ generates the following next-stage Scala code, that is, the
function @x39@:
\begin{lstlisting}
  val x39 = {(x40: List[Set[AbsValue]],          // argument value list
              x41: Map[Addr, Set[AbsValue]],     // store
              x42: Map[Config, Set[ValSt]],      // in cache
              x43: Map[Config, Set[ValSt]]) =>   // out cache
    val x50 = List[Addr](ZCFAAddr("xs"), ZCFAAddr("f")) // addresses of arguments
    val x51 = x50.zip(x40)                              // addresses paired with their values
    val x61 = x51.foldLeft (x41) { case (x52, x53) =>   // prepare a new store that joins
      val x54 = x53._1;  val x55 = x53._2               //   the arguments and their values
        val x58 = x52.getOrElse(x54, x57)               //   into the latest store, intially x41
        val x59 = x58.union(x55)
        x52 + (x54 -> x59)
    } ...
  }
\end{lstlisting}

We borrow the notation traditional modular analysis and realize that
such generated function is a reusable artifact which can be used
modularly.  The generated functions already eliminate the
interpretation overhead and monadic overhead, therefore can be
consider as partial summaries. Although, we still need to run the
generated function at next stage to obtain the analysis result.  At
next stage, we can provide different abstract arugments, store and
caches, depending on the context where it is called, thus the analysis
is modular.

Compared with traditional summary-based analysis, our partial
summaries are represented by executable functions that are
parameterized to the context information. In our approach, the
transition from a whole-program analyzer to modular staged analyzer is
mechanized and for free from staging. We do not need to design any
intermediate representation or data structures of summaries.  In some
scenario, traditional summary-based analyses are able to carry more
information within a mdoule, e.g., the summary of their internal
submodules. The staged approach presented here may still need to run
through the fixed-point for submodules.

\iffalse
\paragraph{Partially-Static Context} Under a polyvariant analysis (like
$k$-CFA), such summary generated by specializing static analyzer even could do
more for calling context by partially-static data optimization. For example,
consider that we have a library consists of three functions @f@, @g@ and @h@,
where @f@ calls @g@ and @h@ internally and returns @y@ fianlly (we extend the
language to have multiple arguments). @g@ and @h@ do not depend on @f@. Now we
would like to modularlly specialize these functions, starting from @f@ with an
initial environment containing @f@, @g@ and @h@.
\begin{lstlisting}
  def f(a, b, c) = val x = g(a, b); val y = h(x, c);  y
\end{lstlisting}

Since $k$ is bounded to a fixed number, after the analysis running $k$ steps
inside of @f@, the calling context will purely depend on the static structure of
our module being analyzed.
\fi

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

To evaluates an assignment, we first evaluate its right-hand side, and then put
the value into the slot where the address of @v@ points to. For simplicity, we
elect to make the value of the assignment be an @void()@ value.
\begin{lstlisting}
case Assign(x, e) =>
  val (v, $\sigma$_*) = ev(e, $\rho$, $\sigma$)
  (void(), put($\sigma$_*, get($\rho$, x), v))
\end{lstlisting}

To evaluates a @while@ loop statement, we evaluate the condition first. Then
similar to how we treat @branch0@, we have a generic @branch@ function but works
on boolean values. For the @true@ branch, we recursively call @ev@ on a newly
constructed expression @Seq(e, While(t, e))@ meaning that first evaluates @e@
and the repeat the loop. Otherwise, for the other branch, we simply return a
void value and current store.
\begin{lstlisting}
case While(t, e) =>
  val (tv, t$\sigma$) = ev(t, $\rho$, $\sigma$)
  branch(tv, ev(Seq(e, While(t, e)), $\rho$, t$\sigma$), (void(), t$\sigma$))
\end{lstlisting}

@@

Interval domain is a relative simple domain which consists of two numeric
fields, an upper bound and a lower bound. Here we annotate the upper bound and
lower bound fields to be of type @Rep[Double]@. The operations such as @+@ of
two intervals produce a new
\begin{lstlisting}
case class Interval(lb: Rep[Double], ub: Rep[Double]) {
  def +(that: Interval): Interval = that match {
    case Interval(lb_, ub_) => Interval(lb + lb_, ub + ub_)
  }
  ...
}
\end{lstlisting}
